Retrofitting Effect Handlers onto OCaml

### 2021-03-25 retro-concurrency.tex
right shepherd :)
14 pages

\begin{figure*}
\begin{minipage}{0.30\linewidth}
  \begin{minipage}{\linewidth}
    \begin{lstlisting}[numbers=left]
external ocaml_to_c
  : unit -> int = "ocaml_to_c"
exception E1
exception E2
let c_to_ocaml () = raise E1
let _ = Callback.register
    "c_to_ocaml" c_to_ocaml
let omain () =
  try (* h1 *)
    try (* h2 *) ocaml_to_c ()
    | with E2 -> 0
  | with E1 -> 42;;
let _ = assert (omain () = 42)
    \end{lstlisting}
    %\vspace{-2mm}
    \subcaption{\texttt{meander.ml}}
    \label{code:meander_ml}
  \end{minipage}
  \begin{minipage}{\linewidth}
    \begin{lstlisting}[language=c,numbers=left]
#include <caml/mlvalues.h>
#include <caml/callback.h>

value ocaml_to_c (value unit) {
  caml_callback(*caml_named_value
    ("c_to_ocaml"), Val_unit);
  return Val_int(0);
}
    \end{lstlisting}
    %\vspace{-4mm}
    \subcaption{\texttt{meander.c}}
    \label{code:meander_c}
  \end{minipage}
\end{minipage}
%
\begin{minipage}{0.30\linewidth}
  \centering
  \includegraphics[scale=0.44]{figures/stock_stack}
  \subcaption{Stack layout before raise \texttt{E1}.}
  \label{fig:stock_stack}
\end{minipage}
%
\begin{minipage}{0.39\linewidth}
\begin{lstlisting}[language=c,basicstyle=\ttfamily\footnotesize]
#0  0x925dc in caml_raise_exn ()
#1  0x6fd3e in camlMeander__c_to_ocaml_83 () at meander.ml:5
#2  0x925a4 in caml_call_ocaml ()
#3  0x8a84a in caml_callback_exn (...) at callback.c:145
#4  caml_callback (...) at callback.c:199
#5  0x76e0a in ocaml_to_c (unit=1) at meander.c:5
#6  0x6fd77 in camlMeander__omain_88 () at meander.ml:10
#7  0x6fe92 in camlMeander__entry () at meander.ml:13
#8  0x6f719 in caml_program ()
#9  0x925a4 in caml_call_ocaml ()
#10 0x92e4c in caml_startup_common (...) at startup_nat.c:162
#11 0x92eab in caml_startup_exn (...) at startup_nat.c:167
#12 caml_startup (...) at startup_nat.c:172
#13 0x6f55c in main (...) at main.c:44
\end{lstlisting}
%\vspace{-2mm}
\subcaption{\texttt{gdb} backtrace before raise \texttt{E1}.}
\label{code:gdb_backtrace}
\end{minipage}
%\vspace{-2mm}
\caption{Program stack on stock OCaml.}
%\vspace{-2mm}
\end{figure*}

@@

\subsection{External calls and callbacks}
\label{sec:external}

@@

\subsection{Exception handlers}
\label{sec:exn_handlers}

@@

\subsection{Stack unwinding}
\label{sec:unwind}

@@

\section{Effect Handlers}
\label{sec:refine}

@@

\subsection{Asynchronous I/O}

@@

\begin{lstlisting}[numbers=left]
let run main =
  let runq = Queue.create () in
  let suspend k = Queue.push (continue k) runq in
  let pending_reads = ref [] in
  let rec run_next () =
    match Queue.pop runq with
    | f -> f ()
    | exception Queue.Empty ->
      match !pending_reads with
      | [] -> () (* no pending reads *)
      | todo ->
        let compl,todo = do_reads todo in
        List.iter (fun (str,k) ->
          Queue.push (fun () -> continue k str) runq) compl;
        pending_reads := todo;
        run_next ()
  in
  let rec spawn f =
    match f () with
    | () -> run_next () (* value case *)
    | effect Yield k -> suspend k; run_next ()
    | effect (Fork f') k -> suspend k; spawn f'
    | effect (In_line ic) k ->
        pending_reads := (ic,k) :: !pending_reads;
        run_next ()
  in
  spawn main
\end{lstlisting}

The scheduler queue |runq| now holds thunks instead of continuations. The value
|pending_reads| maintains a list of pending reads and the associated
continuations (line 4). We handle the |In_line| effect by pushing the pair of
input channel |ic| and continuation |k| to |pending_reads| (line 24), allowing
other threads in the scheduler to run.

@@

\subsection{Static semantics}
\label{sec:static_semantics}

@@

\subsection{Dynamic semantics}

@@

\subsubsection{Syntax} Our expressions (Figure~\ref{sem:syntax}) consist of
integer constants ($n$), variables ($x$), abstraction ($\lam{x}{e}$),
application ($e~e$), arithmetic expressions ($e \odot e$) where $\odot$ ranges
over $\{+,-,*,/\}$, raising exceptions ($\throw{l}{e}$), performing effects
($\perform{l}{e}$), and handling effects ($\handle{e}{h}$). Abstractions come
in two forms: OCaml abstractions ($\lambda^o$) and C abstractions
($\lambda^c$). The handler consists of a return case ($\caseval{x}{e}$), zero
or more exception cases ($\caseexn{l}{x}{e}$) with label $l$, parameter $x$ and
body $e$, and zero or more effect cases ($\caseeff{l}{x}{k}{e}$) with label
$l$, parameter $x$, continuation $k$ and body $e$.

@@

%\vspace{-1mm}
\subsubsection{Top-level reductions}

@@

%\vspace{-1mm}
\subsubsection{C reductions}

@@

%\vspace{-1mm}
\subsubsection{OCaml reductions}

@@

Observe that |continue| and |discontinue| are not part of the expressions. They
are encoded as follows: $\textsf{continue} ~k ~e = (k ~(\olam{x}{x})) ~e$ and
$\textsf{discontinue} ~k ~l ~e = (k ~(\olam{x}{\throw{l}{x}})) ~e$.
Intuitively, resuming a continuation in both the cases involves evaluating the
appropriate abstraction on top of the continuation. We perform the
administrative reductions \textsc{Resume1} and \textsc{Resume2} to evaluate the
arguments to |continue| and |discontinue|. The rule \textsc{Resume} appends the
given continuation to the front of the current continuation, and evaluates the
body of the closure.

\section{Implementation}
\label{sec:impl}

We now present the implementation details of effect handlers in Multicore
OCaml. We assume an x86\_64 architecture for the remainder of this paper.
Although Multicore OCaml currently only supports the x86\_64 backend on Linux,
our design does not preclude other architectures and OSs.

@@

match (* comp_e *)
  match (* comp_f *)
    (*p1*) perform E (*p3*)
	with | v -> v | effect F kf -> ()
with | v -> v
		 | effect E ke -> (*p2*) continue ke ()
    \end{lstlisting}
		\subcaption{Constructing continuation objects}
		\label{code:effimpl}
  \end{minipage}
  %
  \begin{minipage}{0.44\linewidth}
		\vspace{2mm}
    \centering
    \includegraphics[scale=0.42]{figures/fiber_handler}
		\subcaption{Program state for code in~\ref{code:effimpl}}
    \label{fig:fiber_handler}
  \end{minipage}
\end{minipage}
%\vspace{-2mm}
\caption{Layout of Multicore OCaml effect handlers.}
%\vspace{-3mm}
\end{figure*}

@@

%\vspace{-3mm}
\subsection{External calls and callbacks}

@@

%\vspace{-3mm}
\subsection{Effect handlers}
\label{sec:effimpl}

@@

%\vspace{-3mm}
\subsection{Stack unwinding}

@@

Logically, DWARF call-frame information maintains a large table which records
for every machine instruction where the return address and callee-saved
registers are stored. To avoid reifying this large table, DWARF directives
represent the table using a compact bytecode representation. Each block of such
instructions usually describe the call-frame information for a single function,
where each instruction describes the table as a delta of the state at the
previous instruction. While this leads to a compact representation, computing
the return address at a given instruction requires interpreting the bytecode
from the start of the function, which may potentially be very slow.

Within each function, DWARF maintains a \emph{canonical frame address} (CFA)
and is traditionally the stack pointer before entering this function. Hence, on
x86-64, where the return address is pushed on the stack on call, the return
address is at |CFA - 8|. Our goal is to encode the CFA using DWARF directives
when stacks are switched. Recall that stack switching occurs in effect
handlers, external calls and callbacks. When starting to run on a new fiber, we
insert DWARF bytecode to follow the |parent_fiber| pointer and dereference the
|saved_sp| to get the CFA (|saved_sp + 8|). During callbacks into OCaml, we
save the current system stack pointer in the |context block| in
Figure~\ref{fig:fiber} to identify the CFA in the C stack. DWARF unwinding for
external calls is implemented by following a link to the current OCaml stack
pointer. With these changes, we get the same backtrace for the \texttt{meander}
program from \S\ref{sec:unwind}, modulo runtime system functions due to effect
handlers.

We have utilized the DWARF verification tool from~\cite{Bastian19} to flag
possible errors in the hand written DWARF directives where we have altered the
OCaml assembler to implement effect handlers. In three places additional manual
verification was required; two were where the stack is intentionally switched
between fibers and one was where the control returns to the parent fiber which
happened to be resized due to effect handler invocation, which the tool could
not track.

Despite the correct DWARF unwind information, using DWARF to record call stack
information in |perf| only captures the call stack of the current fiber in
Multicore OCaml. Since stack unwinding using DWARF is slow due to bytecode
interpretation overhead, |perf| dumps the (user) call stack when
sampled~\cite{Bastian19}. This only includes the frames from the current fiber.
This is a limitation of |perf| and not of our stack layout. Bastian et
al.~\cite{Bastian19} report on a technique to pre-compile the unwind table to
assembly which speeds up DWARF-based unwinding by upto 25$\times$. With this
fast unwinding, |perf| would no longer needs to dump the call stack, and hence,
we could get the backtrace of the entire call stack and not just the current
fiber.

%\vspace{-2mm}
\subsection{Garbage collection}
\label{sec:gc}

@@

This frees up both the memory allocated for the fiber stack as well as other
system resources, assuming that user code does not handle |Unwind| exception
and fails to re-raise it. But installing a finaliser on every captured
continuation by default introduces significant overhead to continuation capture
(\S\ref{sec:res_final_k}). Hence, we choose not to implement such a finaliser by
default. It is also useful to note that even if the memory for the fiber stack
is managed by the GC rather than being allocated in the C heap as we do now, we
would still need a finalisation mechanism to unwind the stack and release other
system resources that may be held by the continuation.

@@

%\vspace{-2mm}
\section{Evaluation}
\label{sec:eval}

@@

%\vspace{-3mm}
\subsection{No effects benchmarks}

@@

\begin{figure*}
	\includegraphics[scale=0.35]{sandmark-notebook/sandmark_time}
	\caption{Normalized time of macro benchmarks. Baseline is Stock OCaml,
		whose running time in seconds in given in parenthesis.}
	%\vspace{-2mm}
	\label{res:macro_time}
\end{figure*}

\begin{figure*}
	\includegraphics[scale=0.35]{sandmark-notebook/sandmark_codesize}
	\caption{Normalized OCaml text section size (OTSS) of macro benchmarks.
	Baseline is Stock OCaml, whose size in kilobytes given in parenthesis.}
	%\vspace{-2mm}
	\label{res:macro_size}
\end{figure*}

@@

\begin{table}
\caption{Micro benchmarks without effects. Each entry is the percentage
	difference for Multicore OCaml over stock OCaml.}
\vspace{-5mm}
{
\begin{tabular}{r c c c c c c c c c c}
	& \rot{exnval} & \rot{exnraise} & \rot{extcall} & \rot{callback} & \rot{ack}
	& \rot{fib} & \rot{motzkin} & \rot{sudan} & \rot{tak} \\ \hline
	\textbf{Time} & +0.0 & -1.9 & +17 & +65  & +5.3
								& +2.2 & +10 & +0.0 & +4.2 \\
	\textbf{Instr} & +0.0 & +0.0 & +10 & +72 & +16
								 & +24 & +16 & +14 & +17 \\ \hline
\end{tabular}
}
\label{tab:micro_noeffect}
\vspace{-3mm}
\end{table}

@@

The results show that exceptions are no more expensive in |MC| compared to
|stock|. In the other programs, |MC| executes more instructions due to stack
overflow checks. The performance impact on callbacks is more significant than
external calls. For callbacks, since we reuse the current fiber stack, we need
to ensure it has enough room for inserting additional frames, while |stock|
does not need to do this. Callback performance is less important than external
calls, which are far more numerous. As seen previously, these results however
do not affect macro benchmark performance.

%\vspace{-3mm}
\subsection{No perform benchmarks}

@@

As one of our primary goals of adding effect handlers to OCaml is to support
direct style concurrency, we implement a full-fledged HTTP/1.1 web server using
effect handlers (|MC|). We use effect handlers to implement a user-level
threading library that spawns a thread per request. We use
\texttt{httpaf}~\cite{httpaf} for HTTP handling, and
\texttt{libev}~\cite{libev} for the eventloop. We compare our implementation
against an Lwt version (|lwt|) which also uses \texttt{httpaf} and
\texttt{libev}. Unlike using effect handlers, the Lwt version is written in
monadic style and does not have the notion of a thread per request. For
comparison, we include a Go 1.13 version (|go|) that uses the
\texttt{net/http}~\cite{nethttp} package. As both the OCaml versions are single
threaded, the Go benchmark is run with \texttt{GOMAXPROCS=1}.

@@

%\vspace{-2mm}
\section{Related Work}
\label{sec:related}

@@

%\vspace{-3mm}
\section{Conclusions}
\label{sec:conc}

@@

We thank our shepherd, Matthew Fluet, whose comments substantially improved the
presentation. This research was funded via a Royal Commission for the
Exhibition of 1851 and Darwin College Research Fellowships, and by grants from
Jane Street and Tezos Foundation.

>>>>

\begin{figure*}
\begin{minipage}{0.30\linewidth}
  \begin{minipage}{\linewidth}
    \begin{lstlisting}[numbers=left]
external ocaml_to_c
  : unit -> int = "ocaml_to_c"
exception E1
exception E2
let c_to_ocaml () = raise E1
let _ = Callback.register
    "c_to_ocaml" c_to_ocaml
let omain () =
  try (* h1 *)
    try (* h2 *) ocaml_to_c ()
    | with E2 -> 0
  | with E1 -> 42;;
let _ = assert (omain () = 42)
    \end{lstlisting}
    \vspace{-2mm}
    \subcaption{\texttt{meander.ml}}
    \label{code:meander_ml}
  \end{minipage}
  \begin{minipage}{\linewidth}
    \begin{lstlisting}[language=c,numbers=left]
#include <caml/mlvalues.h>
#include <caml/callback.h>

value ocaml_to_c (value unit) {
  caml_callback(*caml_named_value
    ("c_to_ocaml"), Val_unit);
  return Val_int(0);
}
    \end{lstlisting}
    \vspace{-4mm}
    \subcaption{\texttt{meander.c}}
    \label{code:meander_c}
  \end{minipage}
\end{minipage}
%
\begin{minipage}{0.30\linewidth}
  \centering
  \includegraphics[scale=0.44]{figures/stock_stack}
  \subcaption{Stack layout before raise \texttt{E1}.}
  \label{fig:stock_stack}
\end{minipage}
%
\begin{minipage}{0.39\linewidth}
\begin{lstlisting}[language=c,basicstyle=\ttfamily\footnotesize]
#0  0x925dc in caml_raise_exn ()
#1  0x6fd3e in camlMeander__c_to_ocaml_83 () at meander.ml:5
#2  0x925a4 in caml_call_ocaml ()
#3  0x8a84a in caml_callback_exn (...) at callback.c:145
#4  caml_callback (...) at callback.c:199
#5  0x76e0a in ocaml_to_c (unit=1) at meander.c:5
#6  0x6fd77 in camlMeander__omain_88 () at meander.ml:10
#7  0x6fe92 in camlMeander__entry () at meander.ml:13
#8  0x6f719 in caml_program ()
#9  0x925a4 in caml_call_ocaml ()
#10 0x92e4c in caml_startup_common (...) at startup_nat.c:162
#11 0x92eab in caml_startup_exn (...) at startup_nat.c:167
#12 caml_startup (...) at startup_nat.c:172
#13 0x6f55c in main (...) at main.c:44
\end{lstlisting}
%\vspace{-2mm}
\subcaption{\texttt{gdb} backtrace before raise \texttt{E1}.}
\label{code:gdb_backtrace}
\end{minipage}
\vspace{-2mm}
\caption{Program stack on stock OCaml.}
\vspace{-2mm}
\end{figure*}

@@

\vspace{-1mm}
\subsection{External calls and callbacks}
\label{sec:external}

@@

\vspace{-1mm}
\subsection{Exception handlers}
\label{sec:exn_handlers}

@@

\vspace{-1mm}
\subsection{Stack unwinding}
\label{sec:unwind}

@@

\vspace{-1mm}
\section{Effect Handlers}
\label{sec:refine}

@@

\vspace{-1mm}
\subsection{Asynchronous I/O}
\label{sec:async_io}

@@

\begin{lstlisting}[numbers=left]
let run main =
  let runq = Queue.create () in
  let suspend k = Queue.push (continue k) runq in
  let pending_reads = ref [] in
  let rec run_next () =
    match Queue.pop runq with
    | f -> f ()
    | exception Queue.Empty ->
      match !pending_reads with
      | [] -> () (* no pending reads *)
      | todo ->
        let compl,todo = do_reads todo in
        List.iter (fun (str,k) ->
          Queue.push (fun () -> continue k str) runq) compl;
        pending_reads := todo;
        run_next ()
  in
  let rec spawn f =
    match f () with
    | () -> run_next () (* value case *)
    | effect Yield k -> suspend k; run_next ()
    | effect (Fork f') k -> suspend k; spawn f'
    | effect (In_line ic) k ->
      pending_reads := (ic,k)::!pending_reads; run_next ()
  in
  spawn main
\end{lstlisting}

The scheduler queue |runq| now holds thunks instead of continuations. The value
|pending_reads| maintains a list of pending reads and the associated
continuations (line 4). At line 24, we handle the |In_line| effect by pushing
the pair of input channel |ic| and continuation |k| to |pending_reads|,
allowing other threads in the scheduler to run.

@@

\vspace{-1mm}
\subsection{Static semantics}
\label{sec:static_semantics}

@@

\vspace{-1mm}
\subsection{Dynamic semantics}

@@

\vspace{-1mm}
\subsubsection{Syntax}

Our expressions (Figure~\ref{sem:syntax}) consist of integer constants ($n$),
variables ($x$), abstraction ($\lam{x}{e}$), application ($e~e$), arithmetic
expressions ($e \odot e$) where $\odot$ ranges over $\{+,-,*,/\}$, raising
exceptions ($\throw{l}{e}$), performing effects ($\perform{l}{e}$), and
handling effects ($\handle{e}{h}$). Abstractions come in two forms: OCaml
abstractions ($\lambda^o$) and C abstractions ($\lambda^c$). The handler
consists of a return case ($\caseval{x}{e}$), zero or more exception cases
($\caseexn{l}{x}{e}$) with label $l$, parameter $x$ and body $e$, and zero or
more effect cases ($\caseeff{l}{x}{k}{e}$) with label $l$, parameter $x$,
continuation $k$ and body $e$.

@@

\vspace{-1mm}
\subsubsection{Top-level reductions}

@@

\vspace{-1mm}
\subsubsection{C reductions}

@@

\vspace{-1mm}
\subsubsection{OCaml reductions}

@@

Observe that |continue| and |discontinue| are not part of the expressions. They
are encoded as $\textsf{continue} ~k ~e = (k ~(\olam{x}{x})) ~e$ and
$\textsf{discontinue} ~k ~l ~e = (k ~(\olam{x}{\throw{l}{x}})) ~e$.
Intuitively, resuming a continuation in both the cases involves evaluating the
appropriate abstraction on top of the continuation. We perform the
administrative reductions \textsc{Resume1} and \textsc{Resume2} to evaluate the
arguments to |continue| and |discontinue|. The rule \textsc{Resume} appends the
given continuation to the front of the current continuation, and evaluates the
body of the closure.

\vspace{-2mm}
\section{Implementation}
\label{sec:impl}

We now present the implementation details of effect handlers in Multicore
OCaml. While we assume an x86\_64 architecture for the remainder of this paper,
our design does not preclude other architectures and OSs.

@@

match (* comp_e *)
  match (* comp_f *)
    (*p1*) perform E (*p3*)
	with | v -> v | effect F kf -> ()
with | v -> v
		 | effect E ke -> (*p2*) continue ke ()
    \end{lstlisting}
		\subcaption{Constructing continuation objects}
		\label{code:effimpl}
  \end{minipage}
  %
  \begin{minipage}{0.44\linewidth}
		\vspace{2mm}
    \centering
    \includegraphics[scale=0.42]{figures/fiber_handler}
		\subcaption{Program state for code in~\ref{code:effimpl}}
    \label{fig:fiber_handler}
  \end{minipage}
\end{minipage}
\vspace{-2mm}
\caption{Layout of Multicore OCaml effect handlers.}
\vspace{-3mm}
\end{figure*}

@@

\vspace{-2mm}
\subsection{External calls and callbacks}

@@

\vspace{-2mm}
\subsection{Effect handlers}
\label{sec:effimpl}

@@

\vspace{-2mm}
\subsection{Stack unwinding}

@@

Logically, DWARF call-frame information maintains a large table which records
for every machine instruction where the return address and callee-saved
registers are stored. To avoid reifying this large table, DWARF directives
represent the table using a compact bytecode representation. This bytecode must
be interpreted on demand from the start of the function definition to compute
the call-frame information at any given instruction within the function. For
each function, DWARF maintains a \emph{canonical frame address} (CFA) and is
traditionally the stack pointer before entering this function. Hence, on
x86-64, where the return address is pushed on the stack on call, the return
address is at |CFA - 8|.

Our goal is to encode the CFA using DWARF directives when stacks are switched.
Recall that stack switching occurs in effect handlers, external calls and
callbacks. At the entry to a effect handler block, we insert DWARF bytecode to
follow the |parent_fiber| pointer and dereference the |saved_sp| to get the CFA
(|saved_sp + 8|). During callbacks into OCaml, we save the current system stack
pointer in the |context block| in Figure~\ref{fig:fiber} to identify the CFA in
the C stack. DWARF unwinding for external calls is implemented by following a
link to the current OCaml stack pointer. With these changes, we get the same
backtrace for the \texttt{meander} program from \S\ref{sec:unwind}, modulo
runtime system functions due to effect handlers. We have verified the
correctness of our DWARF directives using the verification tool from Bastian et
al.~\cite{Bastian19}.

Despite the correct DWARF unwind information, using DWARF to record call stack
information in |perf| only captures the call stack of the current fiber in
Multicore OCaml. Since stack unwinding using DWARF is slow due to bytecode
interpretation overhead, |perf| dumps the (user) call stack when
sampled~\cite{Bastian19}. This only includes the frames from the current fiber.
This is a limitation of |perf| and not of our stack layout. Bastian et
al.~\cite{Bastian19} report on a technique to pre-compile the unwind table to
assembly, which speeds up DWARF-based unwinding by up to 25$\times$. With this
technique, |perf| can unwind the stack at sample points rather than dumping the
call stack, which captures the complete backtrace rather than just the current
fiber.

\vspace{-2mm}
\subsection{Garbage collection}
\label{sec:gc}

@@

This frees up both the memory allocated for the fiber stack as well as other
system resources, assuming that user code does not handle |Unwind| exception
and fails to re-raise it. Since installing a finaliser on every captured
continuation introduces significant overhead (\S\ref{sec:res_final_k}), we
choose not to do it by default. It is also useful to note that even if the
memory for the fiber stack is managed by the GC, we would still need a
finalisation mechanism to unwind the stack and release other system resources
that may be held by the continuation.

@@

\vspace{-2mm}
\section{Evaluation}
\label{sec:eval}

@@

\vspace{-2mm}
\subsection{No effects benchmarks}

@@

\begin{figure*}
	\includegraphics[scale=0.35]{sandmark-notebook/sandmark_time}
	\vspace{-4mm}
	\caption{Normalized time of macro benchmarks. Baseline is Stock OCaml,
		whose running time in seconds in given in parenthesis.}
	\vspace{-2mm}
	\label{res:macro_time}
\end{figure*}

\begin{figure*}
	\includegraphics[scale=0.35]{sandmark-notebook/sandmark_codesize}
	\vspace{-4mm}
	\caption{Normalized OCaml text section size (OTSS) of macro benchmarks.
	Baseline is Stock OCaml, whose size in kilobytes given in parenthesis.}
	\vspace{-2mm}
	\label{res:macro_size}
\end{figure*}

@@

\begin{table}
\caption{Micro benchmarks without effects. Each entry is the percentage
	difference for Multicore OCaml over stock OCaml.}
\vspace{-5mm}
{
\begin{tabular}{r c c c c c c c c c c}
	& \rot{exnval} & \rot{exnraise} & \rot{extcall} & \rot{callback} & \rot{ack}
	& \rot{fib} & \rot{motzkin} & \rot{sudan} & \rot{tak} \\ \hline
	\textbf{Time} & +0.0 & -1.9 & +17 & +65  & +5.3
								& +2.2 & +10 & +0.0 & +4.2 \\
	\textbf{Instr} & +0.0 & +0.0 & +10 & +72 & +16
								 & +24 & +16 & +14 & +17 \\ \hline
\end{tabular}
}
\label{tab:micro_noeffect}
\vspace{-5mm}
\end{table}

@@

The results show that exceptions are no more expensive in |MC| compared to
|stock|. In the other programs, |MC| executes more instructions due to stack
overflow checks. The performance impact on callbacks is more significant than
external calls. For callbacks, since we reuse the current fiber stack, we need
to ensure it has enough room for inserting additional frames, while |stock|
does not need to do this. Callback performance is less important than external
calls, which are far more numerous.

\vspace{-2mm}
\subsection{No perform benchmarks}

@@

Using effect handlers, we have implemented a full-fledged HTTP/1.1 web server
by extending the example from \S\ref{sec:async_io} (|MC|). The web server
spawns a lightweight thread per request. We use \texttt{httpaf}~\cite{httpaf}
for HTTP handling, and \texttt{libev}~\cite{libev} for the eventloop. We
compare our implementation against an Lwt version (|lwt|) which also uses
\texttt{httpaf} and \texttt{libev}. Unlike using effect handlers, the Lwt
version is written in monadic style and does not have the notion of a thread
per request. For comparison, we include a Go 1.13 version (|go|) that uses the
\texttt{net/http}~\cite{nethttp} package. As both the OCaml versions are single
threaded, the Go benchmark is run with \texttt{GOMAXPROCS=1}.

@@

\vspace{-2mm}
\section{Related Work}
\label{sec:related}

@@

\vspace{-2mm}
\section{Conclusions}
\label{sec:conc}

@@

We thank the PLDI reviewers and our shepherd, Matthew Flatt, whose comments
substantially improved the presentation. This research was funded via a Royal
Commission for the Exhibition of 1851 and Darwin College Research Fellowships,
and by grants from Jane Street and Tezos Foundation.
